---
title: "Refugees in x"
description: ""
author: "Juha Päällysaho"
date: "2023-11-13"
date-modified: "2023-11-20"
execute:
  echo: true
  warning: false
draft: true
---



```{r}
# Option 1: tidytuesdayR package 
## install.packages("tidytuesdayR")
library(tidyverse)
library(gt)
```


```{r}
#| output: false 

tuesdata <- tidytuesdayR::tt_load('2023-08-22')
data_raw <- tuesdata |> pluck(1)
```

```{r}
df <- data_raw |> 
  mutate(year = as_date(parse_date_time(year, "%Y")), .before = 1) |> 
    rename(
        origin_country = coo_name,
        destination_country = coa_name
    )

df
```

```{r}
library(skimr)
df |> skim()
```


# High level

## Time elements


### refugees

* time element of refugees, asymlum seekeers, internally displaced persons (idps) and returned (refugees and idps)
 * these don't have missing data
 


```{r}
refugees_long <- df |> 
    group_by(year) |> 
    summarise(across(where(is.double), \(x) sum(x))) |> 
    pivot_longer(refugees:hst)

```

```{r}
scale_y_mil <- function() {
    scale_y_continuous(labels = scales::label_number(scale = 1/1000000 ,big.mark = " ", suffix = "M"))
}
```


```{r}
refugees_long |> 
    ggplot(aes(year, value, color = name)) +
    geom_point() +
    geom_smooth(method = lm) +
    scale_y_number() +
    labs(title = "how refugees change over time?", subtitle = "linear trend")
```

```{r}
refugees_long |> 
    # filter(!name %in% c("idps", "oip", "ooc") ) |>  
    ggplot(aes(year, value, fill = name)) +
    scale_y_mil() +
    labs(title = "how refugees change over time?",
         subtitle = "per year")
```

```{r}
refugees_long |> 
    filter(str_starts(name, "returned") ) |>
    ggplot(aes(year, value, fill = name)) +
    geom_col() +
    scale_y_continuous(labels = scales::label_number(big.mark = "", decimal.mark = " "))+
    labs(title = "how refugees change over time?",
         subtitle = "per year")
```



### to what countries

```{r}

```


# other focus ideas

* what countries too the most (ratio of gdp)
* from where (coo_name) to destination (coa_name)
* has it changed? why it has changed?

* too many countries, what to focus
  + changes in some region? - why a change has happended?
* short deep dive to asylum seekers?
