---
title: "Refugees Analysis"
description: "Tidytuesday 2023-08-22"
author: "Juha Päällysaho"
date: "2023-11-13"
date-modified: "2023-11-23"
execute:
  echo: true
  warning: false
draft: false
---

# data prep

```{r}
# Option 1: tidytuesdayR package
## install.packages("tidytuesdayR")
library(tidyverse)
```

## read data

```{r}
#| output: false

tuesdata <- tidytuesdayR::tt_load("2023-08-22")
data_raw <- tuesdata |> pluck(1)
```

## skim the data

check data range
missing data from origin_continent
oip and hst are completely missing, ill remove them from dateset
date values are from 2010 to 2022


```{r}
skimr::skim(data_raw)
```


##  continents data

```{r}
countries <- read_csv("https://gist.githubusercontent.com/fogonwater/bc2b98baeb2aa16b5e6fbc1cf3d7d545/raw/6fd2951260d8f171181a45d2f09ee8b2c7767330/countries.csv")

countries_less <- countries |> select(country_code3, continent_name)
```

## enrich 


```{r}
df <- data_raw |>
  select(!c(oip, hst)) |>
  mutate(year = as_date(parse_date_time(year, "%Y")), .before = 1) |>
  rename(
    origin_country = coo_name,
    destination_country = coa_name,
    others_of_concern = ooc
  ) |>
  left_join(countries_less, by = join_by("coo_iso" == "country_code3")) |>
  left_join(countries_less, by = join_by("coa_iso" == "country_code3")) |>
  rename(
    origin_continent = continent_name.x,
    destination_continent = continent_name.y
  )

df |> glimpse()
```

## data to long format

```{r}
group_to_long <- function(.data, ...) {
  .data |>
    group_by(year, ...) |>
    summarise(across(where(is.double), \(x) sum(x))) |>
    pivot_longer(refugees:others_of_concern)
}

refugees_long <- group_to_long(df)
refugees_long_origin <- group_to_long(df, origin_continent)

refugees_long_destination <- group_to_long(df, destination_continent)
```

## get colorspace 
```{r}
library(colorspace)
qualitative_hcl(6)
```

# high level analysis

## notes 

origin and destinations countries which general direction of refugees movement.
refugees group are split to different sub classes, like asylum seekers, internally displaced persons and stateless person 
refugee and internally displaced person also have returned level

## focus

check which refugees group gives interesting analysis layer





## global level

```{r}
scale_y_mil <- function() {
  scale_y_continuous(labels = scales::label_number(scale = 1 / 1000000, big.mark = " ", suffix = "M"))
}

scale_x_yrs <- function() {
  scale_x_date(date_labels = "%y", date_breaks = "4 years")
}


refugees_long |>
  ggplot(aes(year, value, fill = name)) +
  geom_col() +
  scale_y_mil() +
  scale_x_yrs() +
  labs(
    title = "has there been a change in refugees on global level?",
    subtitle = "2010 - 2023"
  ) +
  scale_fill_discrete_qualitative()
```
## continent level

```{r}
plot_continents <- function(.data, .title = "value", .subtitle = "value2", facet_var = origin_continent, .name = name) {
  .data |>
    ggplot(aes(year, value, fill = {{ .name }})) +
    geom_col() +
    scale_x_yrs() +
    scale_y_mil() +
    labs(
      title = .title,
      subtitle = .subtitle
    ) +
    scale_fill_discrete_qualitative() +
    facet_grid(cols = vars({{ facet_var }}))
}

plot_continents(
  refugees_long_origin, "How the refugee level change between the continents?",
  "Origin country of refugees"
)
```

```{r}
plot_continents(
  refugees_long_destination, "How the refugee level change between the continents?",
  "Destination country of refugees", destination_continent
)
```

## refugees between the continets

there seems to be intra continent refugees, next ill check how between the continets and domestic differ

```{r}
df2 <- df |>
  mutate(between_continents = if_else(origin_continent != destination_continent, TRUE, FALSE), .before = 1)
```

```{r}
df2 |>
  filter(between_continents == TRUE) |>
  group_to_long(destination_continent) |>
  plot_continents("what is the destination of Between continents refugees", "", destination_continent)
```
```{r}
df2 |>
  filter(between_continents == TRUE) |>
  group_to_long(origin_continent) |>
  plot_continents(
    "where the Between continents refugees originate from?", "",
    origin_continent
  )
```


# deep dive south america, between continets

south america has interesting change, and I'm not familiar with the continent. I'll try to check if data gives interesting elements

```{r}
southAmerica <- df2 |>
  filter(origin_continent == "South America", between_continents == TRUE)
```

```{r}
southAmericaSummarized <- southAmerica |>
  select(-year) |>
  group_by(origin_country) |>
  summarise(across(where(is.double), list(sum = sum)))


southAmericaSummarized |>
  gt::gt()
```

select refugees and asylum seekers since the other refugree classes are mostly zeros

```{r}
southAmerica |>
  group_to_long(origin_country) |>
  filter(name %in% c("refugees", "asylum_seekers")) |>
  plot_continents("how south america refugees have changed?", "", facet_var = name, .name = origin_country)
```
